The Diagnostic Potential \& Interactive Dynamics of the Colorectal Cancer Virome

### 2018-09-17 doc/manuscript.Rmd
A couple minor updates

The association between the bacterial and viral communities and colorectal cancer was driven by a few important microbes. *Fusobacterium* was the primary driver of the bacterial association with colorectal cancer, which is consistent with its previously described oncogenic potential **(Figure \ref{predmodel} C)**[@Flynn:2016iu]. The virome signature also was driven by a few OVUs, suggesting a role for these viruses in tumorigenesis **(Figure \ref{predmodel} D)**. It is also important to note that while these viruses were driving the signature, the magnitude of their importance and the significance of those values was noticibly less than the bacterial 16S signature, suggesting that unlike what is observed in the bacteria, there are many viruses that are associated with the cancerous state. The identified viruses were bacteriophages, belonging to *Siphoviridae*, *Myoviridae*, and phage taxa who could not be confidently idetnfied beyond their broad phage identification (i.e. "unclassified"). Many of the important viruses, who were confirmed to not have genomic similarity to known bacterial genomes, were unidentifiable (denoted "unknown"). This is common in viromes across habitats; studies have reported as much as 95% of virus sequences belonging to unknown genomic units [@Pedulla:2003tu; @Hannigan:2015fz; @Willner:2009dq; @Brum:2015iaa]. When the bacterial and viral community signatures were combined, both bacterial and viral organisms drove the community association with cancer **(Figure \ref{predmodel} E)**.

@@

## Community Context of Influential Phages
Because the link between colorectal cancer and the virome was driven by bacteriophages (as opposed to non-bacterial viruses), we tested a potential hypothesis that the bacteriophage signal was a mere reflection of the bacterial signal, and thus highly correlated with the bacterial signal. If this hypothesis was true, we would expect a correlation between the relative abundances of influential bacteria and phages. Instead, we observed a strikingly low correlation between bacterial and phage relative abundances **(Figure \ref{correlations} A,C)**. Overall, there was an absence of correlation between the most influential OVUs and bacterial OTUs **(Figure \ref{correlations} B)**. This evidence supported our null hypothesis that the influential phages were not primarily reflections of influential bacteria.

>>>>

The association between the bacterial and viral communities and colorectal cancer was driven by a few important microbes. *Fusobacterium* was the primary driver of the bacterial association with colorectal cancer, which is consistent with its previously described oncogenic potential **(Figure \ref{predmodel} C)**[@Flynn:2016iu]. The virome signature also was driven by a few OVUs, suggesting a role for these viruses in tumorigenesis **(Figure \ref{predmodel} D)**. It is also important to note that while these viruses were driving the signature, the magnitude of their importance and the significance of those values was noticibly less than the bacterial 16S signature, suggesting that unlike what is observed in the bacteria, there are many viruses that are associated with the cancerous state. The identified viruses were bacteriophages, belonging to *Siphoviridae*, *Myoviridae*, and phage taxa that could not be confidently identified beyond their broad phage identification (i.e. "unclassified"). Many of the important viruses, that were confirmed to not have genomic similarity to known bacterial genomes, were unidentifiable (denoted "unknown"). This is common in viromes across habitats; studies have reported as much as 95% of virus sequences belonging to unknown genomic units [@Pedulla:2003tu; @Hannigan:2015fz; @Willner:2009dq; @Brum:2015iaa]. When the bacterial and viral community signatures were combined, both bacterial and viral organisms drove the community association with cancer **(Figure \ref{predmodel} E)**.

@@

## Community Context of Influential Phages
Because the link between colorectal cancer and the virome was driven by bacteriophages (as opposed to non-bacterial viruses), we tested a potential hypothesis that the virome signal was a mere reflection of the bacterial signal, and thus highly correlated with the bacterial signal. If this hypothesis were true, we would expect a correlation between the relative abundances of influential bacterial OTUs and virome OVUs. Instead, we observed a strikingly low correlation between bacterial and viral relative abundances **(Figure \ref{correlations} A,C)**. Overall, there was an absence of correlation between the most influential OVUs and bacterial OTUs **(Figure \ref{correlations} B)**. This evidence supported our null hypothesis that the influential viral OVUs were not primarily reflections of influential bacteria.

### 2018-09-17 doc/manuscript.tex
A couple minor updates

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
    \setmainfont[]{Helvetica}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Diagnostic Potential \& the Interactive Dynamics of the Colorectal Cancer Virome},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\usepackage{setspace}
\doublespacing
\usepackage[vmargin=1in,hmargin=1in]{geometry}
\usepackage{lineno}
\linenumbers
\usepackage{parskip}
\setlength{\parskip}{7.5pt}
\exhyphenpenalty=10000
\hyphenpenalty=10000

@@

  \title{Diagnostic Potential \& the Interactive Dynamics of the Colorectal
Cancer Virome}
    \usepackage{authblk}
    \font\myfont=Helvetica at 12pt
    \font\nextfont=Helvetica at 10pt
                        \author[1]{\myfont Geoffrey D Hannigan}
                    \author[2]{\myfont Melissa B Duhaime}
                    \author[3]{\myfont Mack T Ruffin IV}
                    \author[1]{\myfont Charlie C Koumpouras}
                    \author[1,*]{\myfont Patrick D Schloss}
                            \affil[1]{\nextfont Department of Microbiology \& Immunology, University of Michigan, Ann
Arbor, Michigan, 48109}
                    \affil[2]{\nextfont Department of Ecology and Evolutionary Biology, University of Michigan,
Ann Arbor, Michigan, 48109}
                    \affil[3]{\nextfont Department of Family and Community Medicine, Pennsylvania State
University Hershey Medical Center, Hershey, Pennsylvania, 17033}
                    \affil[*]{\nextfont To whom correspondence may be addressed.}
            \date{}

@@

The association between the bacterial and viral communities and
colorectal cancer was driven by a few important microbes.
\emph{Fusobacterium} was the primary driver of the bacterial association
with colorectal cancer, which is consistent with its previously
described oncogenic potential \textbf{(Figure \ref{predmodel} C)}(22).
The virome signature also was driven by a few OVUs, suggesting a role
for these viruses in tumorigenesis \textbf{(Figure \ref{predmodel} D)}.
The identified viruses were bacteriophages, belonging to
\emph{Siphoviridae}, \emph{Myoviridae}, and phage taxa who could not be
confidently idetnfied beyond their broad phage identification (i.e.
``unclassified''). Many of the important viruses, who were confirmed to
not have genomic similarity to known bacterial genomes, were
unidentifiable (denoted ``unknown''). This is common in viromes across
habitats; studies have reported as much as 95\% of virus sequences
belonging to unknown genomic units (14, 28--30). When the bacterial and
viral community signatures were combined, both bacterial and viral
organisms drove the community association with cancer \textbf{(Figure
\ref{predmodel} E)}.

@@

Differences in the colorectal cancer virome could have been driven by
eukaryotic viruses or by bacteriophages. To better understand the types
of viruses that were important for colorectal cancer, we identified the
virome OVUs as being similar to either eukaryotic viruses or
bacteriophages. The most important viruses to the classification model
were identified as bacteriophages (\textbf{Figure \ref{threewaymodel})}.
Overall, we were able to identify 78.8\% of the OVUs as known viruses,
and 93.8\% of those viral OVUs aligned to bacteriophage reference
genomes. It is important to note that this could have been influenced by
our methodological biases against enveloped viruses (more common of
eukaryotic viruses than bacteriophage), due to chloroform and DNase
treatment for purification.

@@

Because the link between colorectal cancer and the virome was driven by
bacteriophages, we hypothesized that the bacteriophage signal may be a
mere reflection of the bacterial signal, and thus highly correlated with
the bacterial signal. If this hypothesis was true, we would expect a
correlation between the relative abundances of influential bacteria and
phages. Instead, we observed a strikingly low correlation between
bacterial and phage relative abundances \textbf{(Figure
\ref{correlations} A,C)}. Overall, there was an absence of correlation
between the most influential OVUs and bacterial OTUs \textbf{(Figure
\ref{correlations} B)}. This evidence supported our null hypothesis that
the influential phages were not primarily reflections of influential
bacteria, and suggest a more complex relationship between the bacterial
and phage communities.

Given these findings, we hypothesized that the most influential phages
were acting by infecting a wide range of bacteria in the overall
community, instead of just the influential bacteria. In other words, we
hypothesized that the influential bacteriophages were community hubs
(i.e.~central members) within the bacteria and phage interactive
network. We investigated the potential host ranges of all phage OVUs
using a previously developed random forest model that relies on sequence
features to predict which phages infected which bacteria in the
community \textbf{(Figure \ref{network} A)} (34). The predicted
interactions were then used to identify phage community hubs. We
calculated the alpha centrality (i.e.~measure of importance in the
ecological network) of each phage OVU's connection to the rest of the
network. The phages with high centrality values were defined as
community hubs. Next, the centrality of each OVU was compared to its
importance in the colorectal cancer classification model. Phage OVU
centrality was significantly and positively correlated with importance
to the disease model (p-value = 0.004, R = 0.176), suggesting that
phages important in driving colorectal cancer also were more likely to
be community hubs \textbf{(Figure \ref{network} B)}. Together these
findings supported our hypothesis that influential phages were hubs
within their microbial communities and had broad host ranges.

@@

Using our findings here, we have begun a path toward the delineation of
the role the colonic virome plays in colorectal cancer in the form of a
working hypothesis that will guide our future studies \textbf{(Figure
\ref{modelsummary} A)}. We found that basic diversity metrics of alpha
diversity (richness and Shannon diversity) and beta diversity
(Bray-Curtis dissimilarity) were insufficient for identifying virome
community differences between healthy and cancerous states. By
implementing a machine learning approach (random forest classification),
we detected strong associations between the colon virus community
composition and colorectal cancer. The colorectal cancer virome was
composed primarily of bacteriophages. These phage communities were not
exclusively predators of the most influential bacteria, as demonstrated
by the lack of correlation between the abundances of the bacterial and
phage populations. Instead, we identified influential phages as being
community hubs, suggesting phages influence cancer by altering the
greater bacterial community instead of directly modulating the
influential bacteria. Our previous work has shown that modifying colon
bacterial communities alters colorectal cancer progression and tumor
burden in mice (10, 20). This provides a precedent for phages indirectly
influencing colorectal cancer progression by altering the bacterial
community composition. Overall, our data support a model in which the
bacteriophage community modulates the bacterial community, and through
those interactions indirectly influences the bacteria driving colorectal
cancer progression \textbf{(Figure \ref{modelsummary} A)}. Although our
evidence suggested phages indirectly influenced colorectal cancer
development, we were not able to rule out the role of phages directly
interacting with the human host (35, 36).

In addition to modeling the potential connections between virus
communities, bacterial communities, and colorectal cancer, we also used
our data and existing knowledge of phage biology to develop a working
hypothesis for the mechanisms by which this may occur. This was done by
incorporating our findings into the current model for colorectal cancer
development \textbf{(Figure \ref{modelsummary} B)} (22), although it is
important to note that there are also many other alternative hypotheses
by which the system could be operating. We hypothesize that the process
begins with broadly infectious phages in the colon lysing, and thereby
disrupting, the existing bacterial communities. This shift opens novel
niche space that enabled opportunistic bacteria (such as
\emph{Fusobacterium nucleatum}) to colonize. Once the initial
influential founder bacteria establish themselves in the epithelium,
secondary opportunistic bacteria are able to adhere to the founders,
colonize, and establish a biofilm. Phages may play a role in biofilm
dispersal and growth by lysing bacteria within the biofilm, a process
important for effective biofilm growth (37). The oncogenic bacteria may
then be able to transform the epithelial cells and disrupt tight
junctions to infiltrate the epithelium, thereby initiating an
inflammatory immune response. As the adenomatous polyps developed and
progressed towards carcinogenesis, we observed a shift in the phages and
bacteria whose relative abundances were most influential. As the
bacteria enter their oncogenic synergy with the epithelium, we
conjecture that the phages continue mediating biofilm dispersal. This
process would thereby support the colonized oncogenic bacteria by lysing
competing cells and releasing nutrients to other bacteria in the form of
cellular lysates. In addition to highlighting the likely mechanisms by
which the colorectal cancer virome is interacting with the bacterial
communities this model will guide future research investigations of the
role the virome plays colorectal cancer.

At this time, our working hypothesis represents a conceptualization of
areas for future work, which will be required to characterize the
colorectal cancer microbiome at the functional, mechanistic level. There
are many different ways in which this system can be operating, and our
working hypothesis is one. For example, it is possible that the
bacterial communities cause a change in the virome instead of the virome
altering the bacterial communities. Reviewing every possible mechanism
and characteristic of the system is beyond the scope of this manuscript,
but the results we present provide a step toward the answer. Important
future directions will include larger cohort human studies, further
\emph{in vitro} and \emph{in vivo} mechanistic experimentation, and even
attempts at community studies using absolute abundance values instead of
relative abundance, which would allow for more accurate community
dynamic modeling.

As with most microbiome studies it is important to note that, despite
including data for both bacteria and viruses, we are still only studying
a subset of the microbiome. Here we are building on the
bacterial-centric literature around colorectal cancer by beginning to
characterize the virome. The caveat here is that we are technically
limited to studying only the DNA virome, and therefore are not measuring
RNA viruses. This will be an important direction for future studies, as
well as beginning to characterize other additional microbial communities
such as fungi, archaea, etc.

@@

\begin{figure}[htbp]
\centering
\includegraphics{}
\caption{\emph{Basic Quality Control Metrics. A) VLP genomic DNA yield
from all sequenced samples. Each bar represents a sample which is
grouped and colored by its associated disease group. B) Sequence yield
following quality control including quality score filtering and human
decontamination. Dashed line shows rarefaction line in which all samples
with lower sequence yield were excluded from downstream
analysis.}\label{qualcontrol}}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{}
\caption{\emph{Beta-diversity comparing disease states and the study
negative controls which remained following rarefaction. Differences in
means between disease group centroids with 95\% confidence intervals
based on an ANOSIM test with a post hoc multivariate Tukey test.
Comparisons in which the intervals cross the zero mean difference line
(dashed line) were not significantly different.}\label{betaogunegative}}
\end{figure}

>>>>

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
    \setmainfont[]{Helvetica}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={The Diagnostic Potential \& Interactive Dynamics of the Colorectal Cancer Virome},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\usepackage{setspace}
\doublespacing
\usepackage[vmargin=1in,hmargin=1in]{geometry}
\usepackage{lineno}
\linenumbers
\usepackage{parskip}
\setlength{\parskip}{7.5pt}
\exhyphenpenalty=10000
\hyphenpenalty=10000

@@

  \title{The Diagnostic Potential \& Interactive Dynamics of the Colorectal
Cancer Virome}
    \usepackage{authblk}
    \font\myfont=Helvetica at 12pt
    \font\nextfont=Helvetica at 10pt
                        \author[1]{\myfont Geoffrey D Hannigan}
                    \author[2]{\myfont Melissa B Duhaime}
                    \author[3]{\myfont Mack T Ruffin IV}
                    \author[1]{\myfont Charlie C Koumpouras}
                    \author[1,*]{\myfont Patrick D Schloss}
                            \affil[1]{\nextfont Department of Microbiology \& Immunology, University of Michigan, Ann
Arbor, Michigan, 48109}
                    \affil[2]{\nextfont Department of Ecology and Evolutionary Biology, University of Michigan,
Ann Arbor, Michigan, 48109}
                    \affil[3]{\nextfont Department of Family and Community Medicine, Pennsylvania State
University Hershey Medical Center, Hershey, Pennsylvania, 17033}
                    \affil[*]{\nextfont To whom correspondence may be addressed.}
            \date{}

@@

The association between the bacterial and viral communities and
colorectal cancer was driven by a few important microbes.
\emph{Fusobacterium} was the primary driver of the bacterial association
with colorectal cancer, which is consistent with its previously
described oncogenic potential \textbf{(Figure \ref{predmodel} C)}(22).
The virome signature also was driven by a few OVUs, suggesting a role
for these viruses in tumorigenesis \textbf{(Figure \ref{predmodel} D)}.
It is also important to note that while these viruses were driving the
signature, the magnitude of their importance and the significance of
those values was noticibly less than the bacterial 16S signature,
suggesting that unlike what is observed in the bacteria, there are many
viruses that are associated with the cancerous state. The identified
viruses were bacteriophages, belonging to \emph{Siphoviridae},
\emph{Myoviridae}, and phage taxa that could not be confidently
identified beyond their broad phage identification (i.e.
``unclassified''). Many of the important viruses, that were confirmed to
not have genomic similarity to known bacterial genomes, were
unidentifiable (denoted ``unknown''). This is common in viromes across
habitats; studies have reported as much as 95\% of virus sequences
belonging to unknown genomic units (14, 28--30). When the bacterial and
viral community signatures were combined, both bacterial and viral
organisms drove the community association with cancer \textbf{(Figure
\ref{predmodel} E)}.

@@

Differences in the colorectal cancer virome could have been driven by
eukaryotic (human) viruses or by bacteriophages. To better understand
the types of viruses that were important for colorectal cancer, we
identified the virome OVUs as being similar to either eukaryotic viruses
or bacteriophages. The most important viruses to the classification
model were identified as bacteriophages (\textbf{Figure
\ref{threewaymodel})}. Overall, we were able to identify 78.8\% of the
OVUs as known viruses, and 93.8\% of those viral OVUs aligned to
bacteriophage reference genomes. It is important to note that this could
have been influenced by our methodological biases against enveloped
viruses (more common of eukaryotic viruses than bacteriophage), due to
chloroform and DNase treatment for purification.

@@

Because the link between colorectal cancer and the virome was driven by
bacteriophages (as opposed to non-bacterial viruses), we tested a
potential hypothesis that the virome signal was a mere reflection of the
bacterial signal, and thus highly correlated with the bacterial signal.
If this hypothesis were true, we would expect a correlation between the
relative abundances of influential bacterial OTUs and virome OVUs.
Instead, we observed a strikingly low correlation between bacterial and
viral relative abundances \textbf{(Figure \ref{correlations} A,C)}.
Overall, there was an absence of correlation between the most
influential OVUs and bacterial OTUs \textbf{(Figure \ref{correlations}
B)}. This evidence supported our null hypothesis that the influential
viral OVUs were not primarily reflections of influential bacteria.

Given these findings, we posited that the most influential phages were
acting by infecting a wide range of bacteria in the overall community,
instead of just the influential bacteria. In other words, we
hypothesized that the influential bacteriophages were community hubs
(i.e.~central members) within the bacteria and phage interactive
network. We investigated the potential host ranges of all phage OVUs
using a previously developed random forest model that relies on sequence
features to predict which phages infected which bacteria in the
community \textbf{(Figure \ref{network} A)} (34). The predicted
interactions were then used to identify phage community hubs. We
calculated the alpha centrality (i.e.~measure of importance in the
ecological network) of each phage OVU's connection to the rest of the
network. The phages with high centrality values were defined as
community hubs. Next, the centrality of each OVU was compared to its
importance in the colorectal cancer classification model. Phage OVU
centrality was significantly and positively correlated with importance
to the disease model (p-value = 0.004, R = 0.176), suggesting that
phages that were important in driving colorectal cancer also were more
likely to be community hubs \textbf{(Figure \ref{network} B)}. Together
these findings supported our hypothesis that influential phages were
hubs within their microbial communities and had broad host ranges.

@@

Based on our findings, we have developed a conceptual model to be tested
in our future studies aimed at elucidating the role the colonic virome
plays in colorectal cancer \textbf{(Figure \ref{modelsummary} A)}. We
found that basic diversity metrics of alpha diversity (richness and
Shannon diversity) and beta diversity (Bray-Curtis dissimilarity) were
insufficient for identifying virome community differences between
healthy and cancerous states. By implementing a machine learning
approach (random forest classification) to leverage inherent, complex
patterns not detected by diversity measures, we were able to detect
strong associations between the colon virus community composition and
colorectal cancer. The dsDNA virome of colorectal cancer was composed
primarily of bacteriophages. These phage communities were not
exclusively predators of the most influential bacteria, as demonstrated
by the lack of correlation between the abundances of the bacterial and
phage populations. Instead, we identified influential phages as being
community hubs, suggesting phages influence cancer by altering the
greater bacterial community instead of directly modulating the
influential bacteria. Our previous work has shown that modifying colon
bacterial communities alters colorectal cancer progression and tumor
burden in mice (10, 20). This provides a precedent for phages indirectly
influencing colorectal cancer progression by altering the bacterial
community composition. Overall, our data support a model in which the
bacteriophage community modulates the bacterial community, and through
those interactions indirectly influences the bacteria driving colorectal
cancer progression \textbf{(Figure \ref{modelsummary} A)}. Although our
evidence suggested phages indirectly influenced colorectal cancer
development, we were not able to rule out the role of phages directly
interacting with the human host (35, 36).

In addition to modeling the potential connections between virus
communities, bacterial communities, and colorectal cancer, we also used
our data and existing knowledge of phage biology to develop a working
hypothesis for the mechanisms by which this may occur. This was done by
incorporating our findings into a current model for colorectal cancer
development \textbf{(Figure \ref{modelsummary} B)} (22), although it is
important to note that there are also many other alternative hypotheses
by which the system could be operating. We hypothesize that the process
begins with broadly infectious phages in the colon lysing, and thereby
disrupting, the existing bacterial communities. This shift opens novel
niche space that enabled opportunistic bacteria (such as
\emph{Fusobacterium nucleatum}) to colonize. Once the initial
influential founder bacteria establish themselves in the epithelium,
secondary opportunistic bacteria are able to adhere to the founders,
colonize, and establish a biofilm. Phages may play a role in biofilm
dispersal and growth by lysing bacteria within the biofilm, a process
important for effective biofilm growth (37). The oncogenic bacteria may
then be able to transform the epithelial cells and disrupt tight
junctions to infiltrate the epithelium, thereby initiating an
inflammatory immune response. As the adenomatous polyps developed and
progressed towards carcinogenesis, we observed a shift in the phages and
bacteria whose relative abundances were most influential. As the
bacteria enter their oncogenic synergy with the epithelium, we
conjecture that the phages continue mediating biofilm dispersal. This
process would thereby support the colonized oncogenic bacteria by lysing
competing cells and releasing nutrients to other bacteria in the form of
cellular lysates. In addition to highlighting the likely mechanisms by
which the colorectal cancer virome is interacting with the bacterial
communities this model will guide future research investigations of the
role the virome plays colorectal cancer.

Our working hypothesis represents a conceptualization of areas for
future work, which will be required to characterize the colorectal
cancer microbiome at the functional, mechanistic level. There are many
different ways in which this system may operating, and our working
hypothesis is one. For example, it is possible that the bacterial
communities cause a change in the virome instead of the virome altering
the bacterial communities. To better understand this system, future
studies will include larger cohort human studies, further \emph{in
vitro} and \emph{in vivo} mechanistic experimentation, and attempts at
community studies using absolute abundance values instead of relative
abundance, which would allow for more accurate community dynamic
modeling. Overall, this study provides a conceptual foundation to direct
future characterization of the colorectal cancer microbiome at the
functional, mechanistic level.

@@

\begin{figure}[htbp]
\centering
\includegraphics{}
\caption{\emph{Basic Quality Control Metrics. A) VLP genomic DNA yield
from all sequenced samples. Each bar represents a sample which is
grouped and colored by its associated disease group or no-DNA negative
control. B) Sequence yield following quality control including quality
score filtering and human decontamination. Dashed line indicates
rarefaction level (\(10^6\) reads) in which all samples with lower
sequence yields less than this level were excluded from downstream
analysis. After rarefaction and removal of samples with less than
\(10^6\) reads, 27 healthy, 28 cancerous, 27 adenomatous, and 3 negative
control samples remained.}\label{qualcontrol}}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{}
\caption{\emph{Beta-diversity analysis comparing Bray-Curtis
dissimilarity between disease state and negative control community
structures that were captured following sequence rarefaction.
Differences in means between disease group centroids with 95\%
confidence intervals based on an ANOSIM test with a post hoc
multivariate Tukey test. Comparisons in which the intervals cross the
zero mean difference line (dashed line) were not significantly
different.}\label{betaogunegative}}
\end{figure}
